**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-11-25 manuscript/ecosphere_resubmit_2/manuscript.Rmd
second resubmission revisions started

---
title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).   URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).  The entire effort follows rigorous quality control/quality assurance protocols.  These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].  URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event in each of May/June, July/August, and September/October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

Lastly, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in Soranno et al. [-@sorannolagosne2017].

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variable we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we are defining a trend with the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.  Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs]. Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal.

@@

Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).  Also, Rhode Island lakes tended to occur in landscapes (i.e. a 500 meter buffer) with a lower average percent agriculture, more developed land, and comparable forested area than lakes in the LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSENE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].  However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*. Additionally, these results point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.  For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.  Evidence in Rhode Island points to possible lengthening of the growing season as both May and October show increasing temperature trends over time [@noaatrends]. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.  For instance, are oligotrophic lakes showing different trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?  A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a quick, preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.  Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.  The analysis we report in this paper focuses on long term trends at broader scales.  It would be possible to use a similar approach to look at trends of individual lakes which could be a first step in trying to identify drivers of increasing chlorophyll.  These questions are beyond the scope of this study, but all warrant further investigation.

>>>>

---
title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We use the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there are sub-regional (i.e. 3000km^2^) water quality trends. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.  Total phosphorus does show a very slight increase. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, community Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available

@@

The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) which allows for the examination of long-term trends in Rhode Island lakes.   The University of Rhode Island's Watershed Watch (URIWW) is a scientist-led community science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.  Volunteer monitors are trained and then collect field data (e.g., sonde measurements or secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.  The entire effort follows rigorous quality control/quality assurance protocols.  These types of community science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].

URIWW data contributed to the larger regional study by Oliver et al. [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.  The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy. Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.

For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.  In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.  As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.  Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.  The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).

@@

We use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in Soranno et al. [-@sorannolagosne2017].

Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].  Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variables we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

@@

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we evaluated trends based on the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.  Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year or, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs]. Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal.

@@

Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average, than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).  Also, Rhode Island lakes tended to occur in landscapes (i.e. a 500 meter buffer) with a lower average percent agriculture, more developed land, and comparable forested area than lakes in the LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset of surface and epilimnion measurements had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).

@@

Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data. The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.  Panel A. URI Watershed Watch yearly chlorophyll anomalies.  Panel B. LAGOSNE yearly chlorophyll anomalies. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles. Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.

@@

Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].  However, these nutrient reductions may not result in nutrient limitation and when when faced with other changes (e.g. temperature increases) increasing chlorophyll *a* may still occur.  Additionally, our analysis point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.  For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].  Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream]. In short, differences in regional and state level trends are driven by complex and multi-scale processes.

In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.  For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.  Evidence in Rhode Island points to possible lengthening of the growing season as both May and October show increasing temperature trends over time [@noaatrends]. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.  For instance, are oligotrophic lakes showing different trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?  A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.  Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.  The analysis we report in this paper focuses on long term trends at broader scales.  It would be possible to use a similar approach to look at trends of individual lakes which could be a first step in trying to identify drivers of increasing chlorophyll.  These questions are beyond the scope of this study, but all warrant further investigation.
